Nanocall: An Open Source Basecaller for Oxford Nanopore Sequencing Data

### 2016-03-17 paper-biorxiv/paper.tex
fix typo
introduction edits

\abstract{\textbf{Motivation:}
The highly portable Oxford Nanopore MinION sequencer has enabled new applications of genome sequencing directly in the field.  However, the MinION currently relies on a cloud computing platform, Metrichor (\href{metrichor.com}{metrichor.com}), for translating locally generated sequencing data into basecalls.\\
\textbf{Results:}
In this paper we introduce Nanocall, a basecaller for Oxford Nanopore sequencing data as an alternative to Metrichor. Nanocall is freely available, open source and does not require an internet connection to run. However, Nanocall does not replace all of the features of Metrichor - in particular Nanocall does not currently integrate the sequencing of the template and complement strands of the same DNA molecule into a single read (in Metrichor terminology, it produces ``1D'' but not ``2D'' reads). To estimate performace, we ran Nanocall on 4 datasets: \emph{E.coli} and Human, from both natural DNA and PCR-amplified DNA. Nanocall reads have \sim~68\% identity, directly comparable to Metrichor 1D reads, but smaller than \sim~87\% for Metrichor 2D reads. Nanocall took [TODO] $\sim$80hr of single-core CPU time per 100Mbp of 2D data, which is a typical yield for a 24hr sequencing experiment. Nanocall supports multi-threading; on a 4-core CPU, Nanocall would run in time comparable to the sequencer itself.\\
\textbf{Availability:}
Nanocall is available at \href{https://github.com/mateidavid/nanocall}{https://github.com/mateidavid/nanocall} under the MIT license.\\
\textbf{Contact:} \texttt{matei.david at oicr.on.ca}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

The MinION produced by Oxford Nanopore Technologies (ONT) is a highly portable, third-generation sequencing instrument, comparable in size to a standard cell phone. The small form factor makes the MinION particularly suitable for sequencing experiments performed in remote locations~\cite{quick-ebola}. However, the MinION relies on the cloud computing platform \href{metrichor.com}{Metrichor} for \emph{basecalling}, i.e., translating the low-level, locally-generated sequencing data into DNA reads. One can easily envisage a setting in which the scarcity of internet access can limit the effectiveness of using a MinION for sequencing.

Motivated by this situation, we introduce Nanocall, a basecaller for MinION sequencing data. Nanocall does not seek to replace Metrichor, but rather, to provide an offline alternative. In this sense, Nanocall has one important shortcoming compared to Metrichor, in that it performs strand-specific basecalls (``1D'', in Metrichor terminology), but it does not attempt to integrate the information from complementary strands (``2D''). Still, Nanocall might prove useful in situations where internet access is limited, or as a quick pre-Metrichor assessment tool that can be used to inspect basic sequencing characteristics.

In recent developments, ONT have announced a move to a new R9 sequencing pore. All data used in this paper is based on the earlier R7 sequencing pore. ONT have also released Metrichor using a custom Developer License, significantly more restrictive than the MIT License we use for Nanocall.

>>>>

\abstract{\textbf{Motivation:}
The highly portable Oxford Nanopore MinION sequencer has enabled new applications of genome sequencing directly in the field.  However, the MinION currently relies on a cloud computing platform, Metrichor (\href{metrichor.com}{metrichor.com}), for translating locally generated sequencing data into basecalls.\\
\textbf{Results:}
In this paper we introduce Nanocall, a basecaller for Oxford Nanopore sequencing data as an alternative to Metrichor. Nanocall is freely available, open source and does not require an internet connection to run. However, Nanocall does not replace all of the features of Metrichor - in particular Nanocall does not currently integrate the sequencing of the template and complement strands of the same DNA molecule into a single read (in Metrichor terminology, it produces ``1D'' but not ``2D'' reads). To estimate performance, we ran Nanocall on two \emph{E.coli} and two human samples, from both natural DNA and PCR-amplified DNA. Nanocall reads have \sim~68\% identity, directly comparable to Metrichor 1D reads, but smaller than \sim~87\% for Metrichor 2D reads. Nanocall took [TODO] $\sim$80hr of single-core CPU time per 100Mbp of 2D data, which is a typical yield for a 24hr sequencing experiment. Nanocall supports multi-threading; on a 4-core CPU, Nanocall would run in time comparable to the sequencer itself.\\
\textbf{Availability:}
Nanocall is available at \href{https://github.com/mateidavid/nanocall}{https://github.com/mateidavid/nanocall} under the MIT license.\\
\textbf{Contact:} \texttt{matei.david at oicr.on.ca}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

The MinION produced by Oxford Nanopore Technologies (ONT) is a highly portable, third-generation sequencing instrument, comparable in size to a cell phone. The small form factor makes the MinION particularly suitable for sequencing experiments performed in remote locations~\cite{quick-ebola}. However, the MinION relies on the cloud computing platform \href{metrichor.com}{Metrichor} for \emph{basecalling}, i.e., translating the low-level, locally-generated sequencing data into DNA sequence reads. One can easily envisage a setting in which the scarcity of internet access can limit the effectiveness of using a MinION for sequencing.

Motivated by this situation, we introduce Nanocall, a basecaller for MinION sequencing data. Nanocall does not seek to replace Metrichor, but rather, to provide an offline alternative. In this sense, Nanocall has one important shortcoming compared to Metrichor, in that it performs strand-specific basecalls (``1D'', in Metrichor terminology), but it does not attempt to integrate the information from complementary strands (``2D''). Still, we have three major use cases for nanocall: 1) in situations where internet access is limited 2) as a rapid quality assessment check (eg that the correct sample was sequenced) prior to basecalling with Metrichor and 3) as an open source platform for testing new basecalling ideas and models.

ONT has recently announced the development of a new sequencing pore, R9. All data used in this paper is based on the currently available R7 sequencing pore using SQK-MAP006 sequencing kits. ONT has also announced that the source code to their basecaller will be made available under a restrictive proprietary license. We feel an open source alternative will still be a benefit to the development community.
